---
title: "The BCGP Model"
author: "Casey Davis"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(bcgp)
knitr::opts_chunk$set(
  echo = FALSE,
  comment = NA,
  fig.align = "center",
  fig.height = 5,
  fig.width = 8
  )
```

# Introduction - The BCGP Model

The Bayesian Composite Gaussian Process (BCGP) model is meant as a version of
Gaussian Process (GP) regression. It is intended to predict functions 
$y(\mathbf{x}), \mathbf{x} \in \mathcal{X}$, that, when viewed as a draw from a 
stochastic process, exhibit behavior consistent with non-stationarity. The 
"Composite" part of the BCGP model comes from the decomposition of the random 
process, $Y(\mathbf{x})$, into a "global" process, $Y_G(\mathbf{x})$, a "local" 
process, $Y_L(\mathbf{x})$, and an "error" process, $\epsilon(\mathbf{x})$: 
\begin{equation}
  Y(\mathbf{x}) = Y_G(\mathbf{x}) + Y_L(\mathbf{x}) + \epsilon(\mathbf{x})
\end{equation}

The global process is meant to be relatively smooth and to capture the overall 
trend (similar to, say, a polynomial in linear regression). The fact that we 
model the trend with a GP allows us to not have to specify a global trend 
function. The local process is less-smooth and makes local adjustments around the global trend, and the error process accounts for measurement error or acts as a nugget for computational stability.

The model was originally intended to always be composite and non-stationary, and the data was intended to be deterministic output from a computer simulator. The model evolved into having the ability to model data with measurement error, and this R package can fit both stationary and non-stationary and composite and non-composite models and both with and without measurement error. For a non-composite model, the random process is decomposed into 
\begin{equation}
  Y(\mathbf{x}) = Y_R(\mathbf{x}) + \epsilon(\mathbf{x})
\end{equation}

Conditional on model parameters, $\Lambda$, we assume $$ Y(\mathbf{x}) | \Lambda \sim GP\left( \beta_0, C\left(\cdot, \cdot \right)  \right)$$ $\beta_0$ functions as the overall mean, and  $C(\cdot, \cdot)$ is the covariance function. Let $\mathbf{Y} = (y_1, y_2, \ldots, y_n)^\top$ be the observed data, and let $\mathbf{x}$ be the observed data locations in the input space $\mathcal{X}$, a $d$-dimensional finite hyper-rectangle denoted by $[\mathbf{a}, \mathbf{b}]^d = \prod_{j=1}^d[a_j, b_j]$. Let $R(\cdot|\boldsymbol{\rho})$, $\boldsymbol{\rho} = (\rho_1, \ldots, \rho_d)$, $0 \leq \rho_j \leq 1 \;\forall \;j$, be the Gaussian correlation function such that $\mathbf{R}$, an $n \times n$ correlation matrix for the observed data, has elements $$R_{ij} = \prod_{k = 1}^{d}\rho_k^{16\left( x_{ik} - x_{jk} \right)^2}$$

## The Non-Stationary Composite Model

As noted above, the composite GP can be written as the sum of a global process, $Y_G(\mathbf{x})$, a local process, $Y_L(\mathbf{x})$, and an error process, $\epsilon(\mathbf{x})$. The non-stationarity is derived from the fact that the variance of this process is expressed as $\sigma^2(\mathbf{x})$ rather than $\sigma^2$. That is, the variance of the process changes across $\mathcal{X}$, rather than the typical, stationary constant variance. The model parameters are $\Lambda = (\beta_0, w, \boldsymbol{\rho_G}, \boldsymbol{\rho_L}, \sigma^2_\epsilon, \sigma^2(\mathbf{x}), \mu_V, \sigma^2_V, \boldsymbol{\rho_V})^\top$. $\beta_0$ is an overall mean, $w$ is a parameter that weights the global and the local process, $\boldsymbol{\rho_G}$ contains the correlation parameters for the global process, $\boldsymbol{\rho_L}$ contains the correlation parameters for the local process, $\sigma^2_\epsilon$ is the measurement error (or nugget) variance, and $\mu_V, \sigma^2_V$, and  $\boldsymbol{\rho_V}$ are the mean, variance, and correlation parameters for $\sigma^2(\mathbf{x})$, which is itself modeled as a log GP.
 
A typical draw from this process might look like this:
```{r nscExample}
params <- createParameterList()
params$w <- 0.9
params$w <- 0.7
params$rhoG <- 0.7
params$rhoL <- 0.3
params$muV <- -0.3
params$sig2V <- 0.5
params$rhoV <- 0.9
params$sig2eps <- 0
seed <- 701687
set.seed(seed)
nsc <- bcgp:::simulateYGLCompNS(parameters = params)

dataY <- data.frame(x = nsc$x, y = nsc$y)
predY <- data.frame(x = nsc$xPred, y = nsc$yPred)
predG <- data.frame(x = nsc$xPred, y = nsc$yGPred)
predL <- data.frame(x = nsc$xPred, y = nsc$yLPred)
predV <- data.frame(x = nsc$xPred, y = nsc$parameters$VPred)

dataPlot <- ggplot2::ggplot(mapping = ggplot2::aes(x, y)) +
  ggplot2::ggtitle("Non-Stationary Composite BCGP") +
  ggplot2::geom_point(data = dataY, ggplot2::aes(color = "red")) +
  ggplot2::geom_line(data = predY, ggplot2::aes(color = "black")) +
  ggplot2::geom_line(data = predG, ggplot2::aes(color = "blue")) +
  ggplot2::geom_line(data = predL, ggplot2::aes(color = "green")) +
  ggplot2::theme_classic() +
  ggplot2::scale_color_manual(name = NULL,
                              values = c("red" = "red", "black" = "black",
                                         "blue" = "blue", "green" = "green"),
                              labels = c("Observed Data", "True Process", 
                                         "Global Process", "Local Process"),
                              breaks = c("red", "black", "blue", "green"),
                              guide = ggplot2::guide_legend(override.aes = list(
                                linetype = c("blank", "solid", "solid", 
                                             "solid"),
                                shape = c(16, NA, NA, NA)))) +
  ggplot2::ylab("Y(x)") + 
  ggplot2::theme(legend.position = c(0.5, 0.2),
                 plot.title = ggplot2::element_text(size = 16, hjust = 0.5),
                 axis.title = ggplot2::element_text(size = 16)) 

sigmaPlot <- ggplot2::ggplot(mapping = ggplot2::aes(x, y)) +
  ggplot2::ggtitle("Non-Stationary BCGP Example:\nVariance Process") +
  ggplot2::geom_line(data = predV, color = "black") +
  ggplot2::theme_classic() +
  ggplot2::ylab(~ paste(sigma ^ 2, "(x)")) +
  ggplot2::theme(plot.title = ggplot2::element_text(size = 16, hjust = 0.5),
                 axis.title = ggplot2::element_text(size = 16)) 

gridExtra::grid.arrange(dataPlot, sigmaPlot, ncol = 2)

```
Notice that the global process (blue) captures the general trend of the true process (black) while small-scale adjustments around the trend are made by the local process (green). The variance process $\sigma^2(x)$ in this example increases with x, and we can see in the 

## Simulating from a BCGP model
