---
title: "Simulate from the BCGP Model"
author: "Casey Davis"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulate from BCGP Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(bcgp)
knitr::opts_chunk$set(
  echo = TRUE,
  comment = NA,
  fig.align = "center",
  fig.height = 5,
  fig.width = 8
  )
```

## Simulating from (and Plotting) a BCGP model

Simulating from the [Bayesian Composite Gaussian Process](bcgp.html) model is 
straightforward. The function [simulate_from_model](../help/simulate_from_model)
has options that allow the user to simulate from a stationary or non-stationary
model, a composite or non-composite model, deterministic or noisy data, and any
combination of these.

The arguments composite, stationary, and noise determine the type of model the 
process will be generated from and are straightforward.

```{r section1}

nsComp1 <- simulate_from_model(composite = TRUE, stationary = FALSE, noise = FALSE)
z <- plot(nsComp1)

```

Other parameters can control the dimension of the data (d), the number of 
training data locations (n), and the number of test data locations (nTest):

```{r section2}

nsComp2 <- simulate_from_model(composite = TRUE, stationary = FALSE, noise = FALSE, d = 2, n = 25, nTest = 75)
z <- plot(nsComp2)

```

The two simulations above were simulated with randomly generated parameters. The
user can choose the parameters for the model if they wish. 
