---
title: "Simulate from the BCGP Model"
author: "Casey Davis"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulate from BCGP Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(bcgp)
knitr::opts_chunk$set(
  echo = TRUE,
  comment = NA,
  fig.align = "center",
  fig.height = 5,
  fig.width = 8
  )
```

## Simulating from (and Plotting) a BCGP model

Simulating from the [Bayesian Composite Gaussian Process](bcgp.html) model is 
straightforward. The function [simulate_from_model](../help/simulate_from_model)
has options that allow the user to simulate from a stationary or non-stationary
model, a composite or non-composite model, deterministic or noisy data, and any
combination of these. NOTE: the function [bcgpsims](../help/bcgpsims) can be
substituted for `simulate_from_model()`. These two functions create identical S4 objects of class `bcgpsims`. It is the user's preference. 

The arguments `composite`, `stationary`, and `noise` determine the type of model
the process will be generated from and are straightforward.

```{r section1}

nsComp1 <- simulate_from_model(composite = TRUE, stationary = FALSE, 
                               noise = FALSE)
z <- plot(nsComp1)

```

Other parameters can control the dimension of the data (`d`), the number of 
training data locations (`n`), and the number of test data locations (`nTest`):

```{r section2}

nsComp2 <- simulate_from_model(composite = TRUE, stationary = FALSE, 
                               noise = FALSE, d = 2, n = 25, nTest = 75)
z <- plot(nsComp2)

```

The two simulations above were simulated with randomly generated parameters. The
user can choose the parameters for the model if they wish. It is best to first 
use the helper function `create_parameter_list()`, which creates a list of 
parameters, modify that list, and then use that list as the `parameters` 
argument in the `simulate_from_model()` function.

```{r section3}

params <- create_parameter_list(composite = FALSE, stationary = TRUE,
                              noise = FALSE, d = 1)
params
params$rho <- 0.9
params

changeParams <- simulate_from_model(composite = FALSE, stationary = TRUE, 
                               parameters = params)
z <- plot(changeParams, process = "y")

```

The `seed` argument can be set for reproducibility and is straightforward.

If the `decompositon` argument is `TRUE`, then the simulation will return the
global, local, and error processes along with the overall process. It is ignored
if `composite` is `FALSE`.

```{r section4}

decomp <- simulate_from_model(composite = TRUE, stationary = FALSE, 
                              noise = FALSE, d = 1, decomposition = TRUE,
                              seed = 701691)
z <- plot(decomp, process = "y", decomposition = TRUE)

```


The optional argument `randomX1D` is  relevant only with one-dimensional 
training data and determines whether the training data locations should be 
equally spaced or if they should be randomly sampled on $[0, 1]$. 

```{r section5}

params <- create_parameter_list(composite = TRUE, stationary = TRUE,
                              noise = FALSE, d = 1)

randomX <- simulate_from_model(composite = TRUE, stationary = TRUE,
                               randomX1D = TRUE, parameters = params)

seqX <- simulate_from_model(composite = TRUE, stationary = TRUE,
                            randomX1D = FALSE, parameters = params)

p1 <- plot(randomX, process = "y", print = FALSE)
p2 <- plot(seqX, process = "y", print = FALSE)

gridExtra::grid.arrange(p1, p2, ncol = 2)

```

The optional parameter `gridTest` indicates wheter the test data should be 
generated on a grid. For example, if `gridTest = FALSE`, then `nTest` test data
locations are randomly selected on $[0, 1]^d$. However, if `gridTest = TRUE` and
`gridTestSize`, the number of points per dimension, is specified, then the 
$gridTestSize^d$ test data locations are generated on a grid on $[0, 1]^d$ with
`gridTestSize` points for each dimension.

```{r section6}

params <- create_parameter_list(composite = TRUE, stationary = FALSE,
                              noise = FALSE, d = 2)

randomXTest <- simulate_from_model(composite = TRUE, stationary = FALSE, d = 2,
                                  parameters = params, gridTest = FALSE)

gridXTest <- simulate_from_model(composite = TRUE, stationary = FALSE, d = 2,
                                 parameters = params, 
                                 gridTest = TRUE, gridTestSize = 10)

p3 <- plot(randomXTest, process = "y", print = FALSE)
p4 <- plot(gridXTest, process = "y", print = FALSE)

gridExtra::grid.arrange(p3, p4, ncol = 2)

```

One reason for the option for the grid is plotting. If `gridTest = TRUE`, then 
the plot (in 2-D) can be made with the `geom_raster` function in the `ggplot2`
package.

```{r raster}

p5 <- plot(gridXTest, process = "y", print = FALSE, raster = FALSE)
p6 <- plot(gridXTest, process = "y", print = FALSE, raster = TRUE)

gridExtra::grid.arrange(p5, p6, ncol = 2)

```
